---
title: "PPOL670 Assignment 4"
author: "Ruqaya, Min Ha, Parushya"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup, include =FALSE}
# Setting some global options
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
options(kableExtra.auto_format = FALSE)
# Loading packages ----

# check.packages function: install and load multiple R packages.
# Found this function here: https://gist.github.com/smithdanielle/9913897 on 2019/06/17
# Check to see if packages are installed. Install them if they are not, then load them into the R session.
check.packages <- function(pkg) {
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) {
    install.packages(new.pkg, dependencies = TRUE)
  }
  sapply(pkg, require, character.only = TRUE)
}


# Check if packages are installed and loaded:
packages <- c("janitor", "tidyverse","ggthemes",
              "utils", "here","knitr","kableExtra","ggrepel",
              "scales","ggpubr", "tools")
check.packages(packages)

# Theme update
theme_set(theme_bw())
theme_update(plot.margin = unit(c(1,1,1,1), "cm"),
             panel.border = element_blank(),
             legend.position = "bottom",
             plot.caption = element_text(size = 7, face = "italic"))

```

# How has Unemployment Changed Across the World?

Unemployment refers to the share of the labor force that is without work but available for and seeking employment. To answer the question posed above, we look at the unemployment data from across the world in the period 1990 to 2022.[^1].

[^1]: Data sourced from [World Bank Open Data Reository](https://data.worldbank.org/indicator/SL.UEM.TOTL.ZS?view=chart)

```{r load-data}
# This File has unemployment rate for different countries across time
data_unemp <- read_csv("data/API_SL.UEM.TOTL.ZS_DS2_en_csv_v2_4570923/API_SL.UEM.TOTL.ZS_DS2_en_csv_v2_4570923.csv", skip = 3)

# This file has classification according to regions and income groups countries are in today
data_region <- read_csv("data/API_SL.UEM.TOTL.ZS_DS2_en_csv_v2_4570923/Metadata_Country_API_SL.UEM.TOTL.ZS_DS2_en_csv_v2_4570923.csv")

# Merging

working_Data <- data_unemp %>% 
  left_join(
    data_region %>% select(c("Country Code","Region","IncomeGroup")),
    by = "Country Code") %>%  # Joining According to Country Code
  as_tibble() %>% 
  filter( !is.na(Region) # Removing Countries which do not have a Region Assigned
          )

# Converting to Long
working_Data_long <- working_Data %>% 
  pivot_longer(
    cols = starts_with(c("19","20")),
    names_to = "Year",
    values_to = "Unemployment"
  ) %>% 
  select( 
    c("Country Name", "Region","IncomeGroup","Year","Unemployment")
          ) %>% 
   mutate(
    IncomeGroup = factor(IncomeGroup, levels = c("High income", "Upper middle income", "Lower middle income","Low income"))
  )

# Taking Years for which Unemployment NAs are not a big proportion
summary_year <- working_Data_long %>% 
  group_by(Year) %>% 
  summarize(
    n_nas = sum(is.na(Unemployment))
  )
# Post 1990 NAs are just around for 15% of the countries
working_Data_long <- working_Data_long %>% 
  mutate( 
    Year = as.integer(Year)
    ) %>% 
  filter( Year >= 1991 & !is.na(IncomeGroup))
```

# How has World Changed?

```{r world}
working_Data_long %>% 
  ggplot( aes( x = Year, y = Unemployment)) +
  geom_boxplot(aes(group = Year),show.legend = FALSE, outlier.alpha = 0.5, alpha = 0.5) + 
  geom_smooth(method = "lm", color = "blue")+
  labs(x = "Year", y = "Unemployment Rate", title = "Changes Across all Countries in the World",
       caption = "The blue line represents the mean unemployment rate across the period") + 
  theme_minimal() 

```
In this figure, we see that over the years the  aggregatelevels of unemployment has reamined same across the world. However,  the disparity has also reduced slightly as the number of distance of outliers has become less. Since this is a global impact, it could be because of global economic policies on poverty, job creation, and equal access to opportunities, in addition to improved national policies.  


# What About Regions?

```{r region, echo=TRUE}
working_Data_long %>% 
  group_by( Year, Region) %>% 
  summarise(
    mean_unemp = mean(Unemployment, na.rm = TRUE)
  ) %>% 
  ggplot( aes( x =  Year, y = mean_unemp/100, group = Region, color = Region)) +
  geom_point(shape = 4, size = 4) +
  geom_line( linetype = "solid") +
  labs(
    x = "Year", 
    y = "Unemployment Rate", title = "Across Regions", 
    subtitle = "1990-2022",
    color = "World Regions") + 
  theme_minimal() + 
  scale_y_continuous(labels = scales::percent) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(7, "Set2")) +
  theme(
    legend.position = "bottom"
  )

```

This line graph shows the disparity of unemployment rate between regions. Regions such as East Asia, South Asia, and Sub-Saharan Africa have lesser fluctuations over the years, whereas regions such as North Africa, Europe and Central Asia have significant fluctuations. This could be because of shocks in financial cycles that affect particular sectors on which countries in these regions are dependent.  

# Across Countries in Different Income Groups

```{r high-income}
working_Data_long %>% 
  filter( Year == 2021) %>% 
  group_by(Region) %>% 
  mutate( n_countries = n()) %>% 
group_by(IncomeGroup, add = TRUE) %>% 
  summarise(
    perc = n()/first(n_countries)
  ) %>% 
  ggplot( aes( x = Region, y = perc, fill =  IncomeGroup)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(7, "Set2")) +
  theme(
    legend.position = "right"
  ) +
  coord_flip() +
  labs(
    y = "Percent of Countries", 
    x = "Regions", title = "Proportion of Countries by Income Group",
    subtitle = "As coded by World Bank in 2022",
    fill = "Income Group") 

```
When it comes to High Income Countries, North America (100%) and Europe & Central Asia (62%) have the highest rates in comparison to the other regions. Latin America & Caribbeans (42%-43%) comes close to two regions, and ranks 3rd in High Income rates. South Asia (70%) and Sub-Saharan Africa (48%-49%) rank highest in Lower Middle Income, while MENA region and East Asia & Pacific (35%-36%) rank 3rd and 4th in the same category of income. On Low Income, Sub-Saharan Africa ranks highest with 50 percent of its population with high unemployment and poverty rates. Overall, the graph shows that Sub-Saharan Africa has the highest Low Income rates, while North America and Europe & Central Asia have the greatest rank High Income rates.  

```{r fig.height=12, fig.width=20}
 working_Data_long %>% 
  filter( Year %in% c("1991", "2021")) %>% 
  pivot_wider(
    id_cols = c("Country Name", "Region", "IncomeGroup"),
    names_from = Year, values_from = Unemployment
  ) %>% 
  ggplot( aes(x = `1991`/100, y = `2021`/100)) +
  geom_point() +
  facet_grid(IncomeGroup~Region) +
  scale_x_log10(labels = scales::percent) +
  scale_y_log10(labels = scales::percent) +
  geom_abline( alpha = 0.3) +
  labs(
    y = "Unemloyment in 2021", 
    x = "Unemloyment in 1991",
    title = "Change across Income Groups X Regions",
    subtitle = "Data from All Countries") +
  theme( text = element_text(size = 17),
         panel.spacing = unit(2, "lines"))
```
In this particular graph, we have applied the blog base 10 transformation to both the x and y-axis. Also, in comparing the unemployment rate in 1991 and 2021, we have added the 45-degree line to the graph. As the 45-degree line has a slope of one, the upper left of the 45-degree line indicates that the y value(in this case, the unemployment rate in 2021) is higher than the x value(the unemployment rate in 1991). Except for the high-income group in Europe & Central Asia, we can observe that there is a tendency of a rising unemployment rate in 2021 than in 1991. In East Asia, the high-income group has a slightly higher unemployment rate than other income groups. The values gathered around the 45-degree line in Sub-Saharan Africa indicate that the unemployment rate in the region has not gone through dramatic change. Considering the right-top values indicate a high-high unemployment rate, it should be noted that the lower-middle income group in the Middle East & North Africa has had over 10% of the high unemployment rate in both years.  
